\mainmatter

# (PART) Introduction {-}

# What is EMA and why do we need it? {#introduction}

\index{EMA} 
\index{EMA!Synonyms} 
\index{Ambulatory Assessment{EMA!Synonyms}} 
\index{Experience Sampling|see{EMA!Synonyms}} 
\index{Ambulatory Self-reporting|see{EMA!Synonyms}} 
\index{Real-time Data Capturing|see{EMA!Synonyms}} 
\index{Continuous Unified Electronic Diary Method|see{EMA!Synonyms}}
\index{Intensive-longitudinal Study Design|see{EMA!Synonyms}}

Ecological Momentary Assessment (EMA) has many aliases. It is known as 'Experience Sampling' [@Larson1983], 'Ambulatory Assessment' [@EbnerPriemer2009] or 'Ambulatory Self-reporting' [@Conner2012a], 'Real-time Data Capturing', the 'Continuous Unified Electronic Diary Method' [@EllisDavies2012], and as the 'Intensive-longitudinal Study Design' [@Bolger2013]. The different terms stress different aspects of EMA research. All terms, however, refer to research methods that involve the repeated sampling of people's current thoughts, emotions, behaviour, physiological states, and context, in their natural environment, typically (but not necessary) via electronic wearable devices [@Shiffman2008].

EMA has been around for many years. Already in the 1980's, early pioneers used electronic devices to elicit responses from study participants to tap into (mental) health processes in everyday life [see, e.g., @Csikszentmihalyi2014]. Recent years, however, have witnessed a large increase in EMA research. Rapid technological developments, a marked interest in the individual, and a wide recognition of the need to study health-related processes in real-life situations have all contributed to this. 

With the increased adoption of EMA, comes a growing need for methodological guidelines. EMA studies have unique characteristics that require specialised research skills, related to study design and statistical analysis. In many cases, these skills are not part of the standard curriculum of academic departments. This book was written to fill this gap.

## What is EMA?

### Self-report versus Observational EMA
\index{EMA!Self-report EMA} \index{EMA!Active EMA|see{Self-report EMA}}
\index{EMA!Observational EMA} \index{EMA!Passive EMA|see{Observational EMA}}

In EMA research, there are two forms of data collection: EMA based on self-report (active) and EMA based on observational (passive) data. Self-report requires participants to provide information, for example by rating their current mood, activities or social interaction on items that are presented to them during EMA. For observational EMA, information is collected through wearables or logfiles without active involvement of participants, for example on heart-rate, activity, smartphone use or engagement on social media. Some studies combine both forms of data collection, for example by using both a self-report sleep diary and an accelerometer in order to study sleeping patterns [@Meijden2016]. 


### EMA sampling
\index{Sampling!Signal-contingent}
\index{Sampling!Event-contingent} 
\index{Sample plan|see{Sampling}}

EMA sampling may focus on a single time-point (signal-contingent sampling), an event (event-contingent sampling), or a combination of both [@Conner2012b]. 

In *signal-contingent sampling*, participants respond to questions when they are prompted to do so by a signal.  Signal-contingent sampling can follow a fixed or a random scheme. In a fixed scheme, participants are prompted at fixed time-points, for example at 9.30, 12:30, and 16:30. In a random scheme, prompts are sent at random time points, typically in pre-set intervals, for example, participants could be prompted to complete two assessments per day, one at a random time point between 10:00 and 14:00, and one at a random time point between 14:00 and 16:00. Using pre-set intervals ensures that participants do not receive several prompts within a limited time-frame [@Piasecki2007]. In addition, it ensures that participants are not bothered by prompts at inappropriate times (e.g., most participants do not appreciate prompts after 22:00 and before 7:30).

In *event-contingent sampling*, study participants complete an assessment whenever a specific event occurs, such as a panic attack or alcohol consumption. One option is to simply instruct the participants to do so. In that case, it is important to be clear, in the instructions, to provide a clear definition of the target event, and to stress the importance to rate each event. In some cases, it may be possible to trigger event-based prompts automatically, for example by linking a self-report EMA questionnaire to an automatically detected change in activity level [@Smyth2003].


## Why EMA?
\index{Recall bias}
\index{Ecological validity} 
\index{Microprocesses}

EMA aims to "minimize recall bias, maximize ecological validity, and allow study of microprocesses that influence behavior in real-world contexts." [@Shiffman2008]. 
Recall bias refers to the concern that in retrospective self-report information, participants' current state might affect the way they remember past events or emotions [@Shiffman2008; @Moore2016]. The idea of EMA is to evaluate participants' *current state*, rather than reflect on past experiences. By making sure participants are measured in their own environment (daily life) instead of a research specific context, such as a lab or mental healthcare centre, EMA is thought to provide *ecological valid* information that is not distorted by context.  

###Focus on the Individual
\index{Nomothetic research} \index{Ideographic research} 
\index{Intra-individual process|see{Ideographic research}}

EMA offers a unique opportunity to study change in individuals with repeated sampling. By doing this, EMA offers a quantitative method for idiographic research [@allport1937], measuring characteristics of (unique) individuals across time and context (within-subject) [@Shiffman2008]. This contrasts with more classic, nomothetic research, in which individuals are grouped together in order to study universal characteristics (between-subject). The relevance of idiographic research is illustrated in Figure 1 (\@ref(fig:fig01a)), where the individual effects of x on y are positive for individuals, but negative for the group as a whole. 
In contrast to more qualitative idiographic studies, such as interviews and N=1 case studies, EMA leads to a large amount of information on each individual. This allows for testing of hypotheses. Also, by studying patterns in individuals (the intra-individual process), you can test whether individuals have similar relations between variables and which factors account for variability [@Conner2009]. This, then, allows you to make inferences to a larger population, combining both idiographic and nomothetic approaches.  

```{r fig01a, echo = FALSE, fig.cap = "An extreme case of how individual effects may be different from group effects: the effect of x on y is positive for individuals (marked by ellipses), but negative for the group (as illustrated by the negative regression line)."}

# simulate individual
sim_p <- function(id = 1, intercept = 0, x_start = 0){
  X <- (0:9)
  Y <- intercept + X * .9 + rnorm(10, 0, 1.5)
  data.frame(id = id, X = X + x_start, Y = Y)
}

# define effect of x on y for the group
intercept <- 100
X <- seq(1, 100, 10)
Y <- 100 - .4 * X 

# let individuals define the group
dt <- NULL
for (i in 1:10) {
  p <- sim_p(i, Y[i], X[i])
  dt <- rbind(dt, p)
}

# plot figure
ggplot(dt, aes(x = X, y = Y, group = id)) + 
  geom_smooth(aes(group = NULL), 
              color = "black", 
              method = "lm", 
              se = FALSE) +
  geom_point(size = 1.2, shape = 16) + 
  stat_ellipse(color = "grey", level = .99, type = "t") + 
  theme_bw(12) + 
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())
```


### Change over time

In more traditional longitudinal research, individuals are often assessed a limited amount of times. For example, by administering a baseline assessment, and 3-month, 6-month and 12-month follow-up. This will provide insight on a macro-level how groups behave over time. Conversely, EMA focuses on micro-processes by measuring multiple times per day for two weeks (\@ref(fig:fig01b)). 

```{r fig01b , echo = FALSE, fig.cap = "34 weeks of mood data, from a single participant"}
# plot figure
ggplot(csd, aes(x = date, y = as.numeric(mood_down))) +
  geom_smooth(method = "loess", span = .05, se = FALSE, colour="black") + 
  geom_point(size = .3, alpha = .3, position = position_jitter(height = .1),
  colour="black") +  scale_x_date() + scale_y_continuous(breaks = 1:7) +
  xlab("Time") + ylab("Mood") +
   theme(panel.background = element_blank())
```  


### Real-world context and ecological validity
\index{Exposomics}
A key feature of EMA is the capture of data in real-world environments as subjects go about their daily activities, as opposed to data collection within a controlled lab or research setting [@Shiffman2008]. Thus, EMA data provides ecological validity and can better generalize to the subject’s lived experience or real world. Practical applications derived from the data would be more relevant and useful to life situations.  


  
## EMA research
In recent decades, research on mental health using EMA has produce an impressive trove of findings that have supported and sometimes, challenged existing theories on behaviour. A key implication of EMA data is the temporal fluidity of cognition, affect, and behaviour which are highly responsive to the context [@Shiffman2009]. EMA, whether collected as self-report or via wearable device/sensor, can have diagnostic, monitoring, management, or intervention applications (@Patel2012; @Aung2017; @Evenson2015]. Below is a non-exhaustive summary of systematic reviews or meta-analyses of research on mental health using EMA.

### Self-report EMA
\index{Self-report EMA}

Mood disorders have been well-studied using EMA methods [@Wenze2010] with several reviews outlining recent findings in depression [@Telford2012; @Wichers2011; @burton2013], anxiety disorders [@Walz2014], depression/bipolar disorder [@AanhetRot2012].
EMA is particularly suited to examine the role of emotions in development and maintenance of obesity and eating disorders [@Engel2016]. Meta-analytic results suggest that negative affect, rather than hunger, is associated with binge eating among individuals with eating disorders [@HaedtMatt2011; @Haedt2012]. 
EMA methods have significantly contributed to the understanding of the processes that drive substance use, cessation, and relapse, often in contrast with theory-driven studies largely derived from global reports collected through questionnaires [@Shiffman2009].

### Passive EMA
\index{Passive sensing|see{EMA!Observational EMA}}
\index{Passive EMA|see{EMA!Observational EMA}}

A systematic review and meta-analysis on actigraphy reported diurnal variations in activity levels among individuals with depression [@burton2013]. The potential of passive sensing via smartphone for mental health research are outlined in two reviews, encompassing the assessment of health and well-being [@Cornet2017], and the measuring, understanding, and intervening in mental illness and maintaining mental health [@Aung2017]. Compared with traditional self-reports, passive sensing is less intrusive and provides more accurate data, continual monitoring, and feedback. However, studies identified in that review did not include clinical samples.

EMA has potential to make significant contributions to mental health research. However, there are some general limitations to consider.
\index{Limitations} 
\index{Limitations!Selected samples}
\index{Limitations!Practive effect}
\index{Limitations!Feasibility}
\index{Limitations!Privacy}

-	Selected samples, e.g. hospital in-patients [@burton2013]
-	Practice effect [@Telford2012], reactivity and compliance [@Shiffman2009]
-	Issues of feasibility and tolerability of prolonged and intense data periods of data collection [@Wichers2011]
-	Issues of privacy, consent, and awareness [@Cornet2017]
-	Responsibility of researcher, e.g. in suicide ideation research [@Wenze2010]


Table: (\#tab:emareviews) Reviews of EMA studies targeting specific Mental
Health Conditions

| **Topic**        | **Author (Year)**        | **Outcome**        |
|:--------------------|:--------------------|:--------------------|
| Anxiety disorders | | |
| | @Walz2014 | Provides insights to the temporal variability of symptoms, and associations between daily affect, behaviours, and situational cues. Combination of EMA and ambulatory assessment of physiological variables and treatment evaluations |
| | | |
| Eating disorders | | |
| | @Engel2016| An overview of studies on eating disorders, obesity, and bariatric surgery using EMA |
| | | |
| Mood disorders | | |
| | @AanhetRot2012 | Provides an overview of EMA studies on correlates of mood, treatment effects, residual symptoms of remitted patients, paediatric populations, MDD/BD specificity, and links with neuroscience |
| | @Aung2017 | Provides conceptual review of passive sensing techniques for measuring, understanding, and treatment of mental illness, and maintenance of mental health |
|| @burton2013 | Diurnal variations in activity levels among depressed individuals |
| |@Telford2012 | Identified six themes of EMA research in MDD: methodology; positive and negative affect; cortisol secretion; antidepressant treatment; work performance; genetic risk factors |
| | @Wenze2010 | Provides an overview of EMA in mood disorder research comprising techniques used, types of population assessed, types of research questions, and a discussion of the potential of EMA in treatment setting of mood disorders |
| | @Wichers2011 | Provides an overview for the possible clinical application of EMA in the diagnostic and treatment of MDD |
| | | |
| Substance abuse| | |
| | @Shiffman2009 | Review of processes that drive substance use, cessation, and relapse, sometimes in contrast with theory-driven studies that are largely derived from global reports collected through questionnaires |
| | | |
| Mental health/ Well-being | | |
| | @Cornet2017 | Outlines the potential of passive sensing to detect status change and behaviour change following feedback on behaviour. Also outlines the challenges of using passive sensing |