---
editor_options:
  chunk_output_type: console
output:
  html_document: default
  pdf_document: default
---

# R packages for EMA research {#rcat}

Many R packages exist that can help you in the management and analysis of EMA data. In this chapter, a selection of these packages are discussed. For each, we provide a summary description, a code example code, and pointers to further documentation, to give you a head start in using the packages for your work.

Table: (\#tab:rcat) List of R packages that are useful in EMA research.

+-----------------+-------------------+-----------------------------------------+
| Category        | Package           | Description                             |
+:================+:==================+:========================================+
| Accelerometry   | GENEAread         | Import GENEActive data into R           |
+-----------------+-------------------+-----------------------------------------+
|                 | GGIR              | Pre-proces and analyse raw multi-day    |
|                 |                   | multi-day accelerometer data.           | 
+-----------------+-------------------+-----------------------------------------+
|                 | PhysicalActivity  | Analyse Actigraph accelerometer data.   | 
+-----------------+-------------------+-----------------------------------------+
| Data Management | dplyr             | Data transformation                     | 
| & Visual        |                   |                                         |
| Exploration     |                   |                                         |
+-----------------+-------------------+-----------------------------------------+
|                 | ggplot2           | Create graphs                           |
+-----------------+-------------------+-----------------------------------------+
|                 | haven             | Import and export SPSS data files       |
+-----------------+-------------------+-----------------------------------------+
|                 | lubridate         | Manipulate date and time variables      |
+-----------------+-------------------+-----------------------------------------+
| Mixed-effects   | lme4              | Fit linear and nonlinear mixed-effects  |
| Modeling        |                   | models. Fast alternative to             |
|                 |                   | package 'nlme'.                         |
+-----------------+-------------------+-----------------------------------------+
|                 | nlme              | Fit linear and nonlinear mixed effects  |
|                 |                   | models. Pre-dates package 'lme4', but is|
|                 |                   | still used because it a provides more   |
|                 |                   | advanced options to model correlational |
|                 |                   | structures in the data.                 |
+-----------------+-------------------+-----------------------------------------+
| Power Analysis  | simr              | Simulation-based power calculations for |
|                 |                   | mixed models.                           | 
+-----------------+-------------------+-----------------------------------------+
| Simulation      | simstudy          | Simulate study data.                    | 
+-----------------+-------------------+-----------------------------------------+
| Spatio-temporal | adehabitatHR      | Developed for home range estimation of  |
| analysis        |                   | wild animals from GPS data. Useful for  |
|                 |                   | human data as well (see Chapter         |
|                 |                   | \@ref(homerange)).                      | 
+-----------------+-------------------+-----------------------------------------+
| Symptom Networks| autovarCore       | Automate the construction of vector     |     
|                 |                   | autoregressive models.                  |
+-----------------+-------------------+-----------------------------------------+
|                 | bootnet           | Assess the stability of symptom         |
|                 |                   | networks.                               |
+-----------------+-------------------+-----------------------------------------+
|                 | qgraph            | Estimate and plot symptom networks.     |  
+-----------------+-------------------+-----------------------------------------+
| Time series     | lomb              | Calculate the Lomb-Scargle Periodogram  |
| analysis        |                   | for unevenly sampled time series.       |
+-----------------+-------------------+-----------------------------------------+


## Accelerometry

Accelerometer data need considerable pre-processing before final analyses can be run. Raw data have to be read in from a variety of brand-specific file formats, data have to re-calibrated on a per-device basis, non-wear periods have to detected, and summarizing measures, such as activity counts and engergy-expenditure measures, have to be calculated from imputed triangular (x, y, x) data, often in several time windows (i.e., epochs). 

### Package GENEAread

GENEActive, sold by Activinsights, is a wrist-worn tri-axial accelerometer that is often used in clinical research studies. With package ***GENEAread*** [@R_GENEAread], raw data can be imported into R for further processing, as illustrated below.    

```{r out.with = "100%", fig.asp = .4, fig.cap = "Raw sensor data of a GENEActive wrist-worn tri-axial accelerometer (down-sampled from 100Hz to 5Hz).", message = FALSE, warning = FALSE}
library(GENEAread)
library(tidyr)

dat <- read.bin(system.file("binfile/TESTfile.bin", package = "GENEAread"),
                verbose = FALSE, downsample = 20)

d <- as.data.frame(dat$data.out)
d <- gather(d, key = "sensor", value = "value", -timestamp)
d$timestamp <- as.POSIXct(d$timestamp, 
                          origin = "1970-01-01",
                          tz = "UTC")

ggplot(d, aes(x = timestamp, y = value)) + 
  geom_line() + 
  facet_wrap(~sensor, scales = "free_y")
```


### Package 'GGIR'
\index{Packages!GGIR}

Package *GGIR* [@R-GGIR] is a packages to pre-process raw acceleromety data from three brands of wearables that are widely used in sleep and physical activity research: [GENEActiv](https://www.geneactiv.org/), [ActiGraph](http://actigraphcorp.com/) and [Axivity](http://axivity.com/).


### Package 'PhysicalActivity'
\index{Packages!PhysicalActivity}

Package ***'PhysicalActivity'*** [@R-PhysicalActivity] provides an alternative to package 'GGIR', when ActiGraph data are available.

```{r fig11a, fig.cap="Activity Counts (5-minute windows), in a Three-day Accelerometer data set.", out.width = "100%", fig.asp = .4}
library(PhysicalActivity)
library(ggplot2)

data(dataSec)

d <- dataCollapser(dataSec, TS = "TimeStamp", col = "counts", by = 300)

ggplot(d, aes(x = as.POSIXct(TimeStamp), y = counts)) +
  geom_line(size = .5, alpha = .5) +
  xlab("Time") + ylab("Activity Counts")
```

## Data management & Visual Exploration
\index{Packages!tidyverse}

The tidyverse is a collection of well-designed packages, authored by the team behind RStudio, that together add a consistent, modern, and efficient extension of base R functionalities. The tidyverse includes popular packages such as "ggplot2" (for plotting), "haven" (to read SPSS files), "dplyr" (for data manipulation), and many more (see: http://tidyverse.org for a full list).

### Package `dplyr`
\index{Packages!dplyr}

Package ***'dplyr'*** [@R-dplyr] implements the 'split-apply-combine'-strategy. With 'dplyr', elementary data manipulations can be e chained (using the pipe operator '%>%') to elegantly implement complex data transformations. 

```{r cs18c}
# code snippet 18.3: aggregate data by ID, through a 'pipe'
require(dplyr)

d <- data.frame(
  c = factor(rep(1:5, each = 10)), 
  score = rnorm(50)
)

b <- as_tibble(d) %>% 
  group_by(c) %>%
  summarize(mean_score = mean(score))
```

A good introduction to dplyr can be found in the book 'R for Data Science' [@wickham2016r], which can be freely accessed online (http://r4ds.had.co.nz/).

### Package 'ggplot2'
\index{Packages!ggplot2}

Package ***'ggplot2*** [@wickham2016ggplot2] provides a collection of high-level plotting commands with which graphs can be build up in layers. It is based on 'The Grammar of Graphics' [@wilkinson2006], an influential analysis of the structure of scientific graphs. Systematic introductions are available on the [the tidyverse website](http://ggplot2.tidyverse.org/), and in the book 'ggplot2: Elegant Graphics for Data Analysis' [@wickham2016ggplot2].

The example below illustrates how graphs are layered. In the first step, a coordinate system is set up. In step 2, all time/scores points are plotted. In step 3, a smoothed line is fitted through these points. Finally, in step 4, the graph is split on a variable ID (a subject identifier), to show individual trajectories.

```{r cs18b, fig.cap = "Plotting layers with ggplot2", out.width = '45%', fig.asp = .4, fig.show = "hold"}
# code snippet 18.2: simple ggplot example
library(ggplot2)

# simulate example data 
d = data.frame(
  ID    = rep(1:4, each = 25),
  time  = rep(1:25, 4), 
  score = rnorm(100, 0, 2))

# step 1: initialise the coordinate system
g <- ggplot(d, aes(x = time, y = score)); g

# step 2: add scatterplot
g <- g + geom_point(); g

# step 3: fit a smoothed line
g <- g + geom_smooth(); g

# step 4: split plot by ID
g + facet_wrap(~ ID)
```


### Package 'haven'
\index{Packages!haven}
\index{SPSS}

With package ***haven*** [@R-haven], SPSS, STATA and SAS files can be read into R.

```{r}
library(haven)

path <- system.file("examples", "iris.sav", package = "haven")
d <- read_sav(path)

attributes(d$Species)
```

### Package 'lubridate'
\index{datetime variables}
\index{Packages!lubridate}

EMA data analyses frequently require manipulations of datetime variables. For this, package ***lubridate*** [@R-lubridate], which provides many functions for common date and datetime operations,  can be very useful. 

In the code snippet below, for example, the 'round_date' function is used to calculate the ENMO value from raw tri-axial accelerometer data, in 15-minute epoch windows.

```{r}
library(emaph)
library(lubridate)
library(ggplot2)
library(dplyr)

d <-  subset(genea,
              timestamp > "2018-06-02 12:00" &
              timestamp < "2018-06-02 18:00")
d$epoch <- round_date(d$timestamp, "minute")

d <- d %>% group_by(id, epoch) %>% 
  summarise(svm = sum(sqrt(x^2 + y^2 + z^2) -1) / length(x))
```

```{r echo = FALSE}
knitr::kable(head(d))
```

To learn more about handling dates and times with 'lubridate', [Chapter 16](http://r4ds.had.co.nz/dates-and-times.html) of the book 'R for Data Science' [@wickham2016r] provides a good introduction.


## Mixed-effects modeling

Several R-packages for mixed-effects modelling exist. The most popular are package **nlme** [@R-nlme] and package ***lme4** [@Bates2015]. Both packages are actively used, as both provide unique functionalities. 

### nlme
\index{Packages!nlme}

Package **nlme** is introduced in chapter \@ref(lmm). It is an older package (in comparison to package lme4), that is still used a lot because it provides options to model correlational structures that are not implemented (yet) in lme4.

```{r cs18d, nlme_example}
# code snippet 18.4: fit a linear mixed model, with lme
library(nlme)
fm <- lme(distance ~ age + Sex, data = Orthodont, random = ~ 1)

fixef(fm)
```

### lme4
\index{Packages!lme4}

Package **lme4** [@Bates2015] is a faster R-reimplementation of the mixed-effects model. With large data sets and complex hierarchical models, this package should probably be preferred. As can be seen below, models specifications in 'lmer' are different from model specifications in 'lme'. 

```{r cs18e, lme4_example, eval = TRUE}
# code snippet 18.4: fit a linear mixed model, with lme
library(lme4)
fm <- lmer(distance ~ age + Sex + (1 | Subject), data = Orthodont)
fixef(fm)
```

## Power analysis

### Package 'simr'
\index{Packages!simr}

With package 'simr' [@Green2016], power of mixed-effects models can be determined via simulation. As illustrated below, the procedure requires the researcher to define the "true" parameters of a mixed model, and a single data set. Then,  function 'simPower' can be used to simulate new datasets and tests (of a specified parameter in the model), to determine the power of the test. 

```{r cs3c}
# code snippet 3.3: Power analysis of a two-group repeated measures design
# (simulation approach)
library(simr)

# construct design matrix 
t <- 1:24
s <- 1:40 
X   <- expand.grid(t = t, s = s)
X$g <- c(rep(0, 24), rep(1, 24))

# fixed intercept and slope
b <- c(2, -0.1, 0, -0.5)

# random intercept variance
V1 <- 0.5                     

# random intercept and slope variance-covariance matrix
V2 <- matrix(c(0.5, 0.05, 0.05, 0.1), 2) 

# residual standard deviation
s <- 1 

model1 <- makeLmer(y ~ t * g + (1 + t | s), 
                   fixef = b, 
                   VarCorr = V2, 
                   sigma = s, 
                   data = X)

powerSim(model1, 
         fixed("t:g", "lr"), 
         nsim = 10,
         progress = FALSE)
```

## Spatio-temporal analysis: 

### GPS data: Package `adehabitatHR`
\index{Packages!adehabitatHR}

Package **adehabitatHR** [@Calenge2006] was created for the study of the habitat of wild animals, using accelerometer and GPS data. Defined procedures can be used for the analysis of human data as well.

```{r cs18g, eval = FALSE}
library(adehabitatHR)


```


## Symptom Network Analysis
\index{symptom networks}

When EMA is used to tap various symptoms, network analysis can reveal the dynamic interplay between these symptoms (see Chapter \@ref(networks). Various packages exist to fit these networks in R. With these packages, it is relatively easy to fit a graphical network on multivariate data sets. If you are interested in conducting a network analysis, be sure to visit the Psycho-systems website, at http://psychosystems.org. 

### Package 'autovarCore`
\index{Packages!autovarCore}
\index{Packages!autovar}

\index{vector autoregressive models}

Vector autoregressive (VAR) models can be used to detect lagged relationships between multiple timeseries. In VAR, each variable is modelled as a linear function of past values (lags) of itself and of present and past values of other variables. When EMA is used to capture multiple phenomena over time, VAR can provide insight in how these phenomena interact. One challenge in VAR modelling is that many alternative models potentially exist, Package ***'autovarCore'*** [@R-autovarCore] was developed to help researchers to find the VAR model with the best fit to a given timeseries data set.

In the (unrealistic) example below, function 'autovar' is used to detect that changes in depression are positively related to past (lag 1) values of activity, in a simulated data set: 

```{r eval = TRUE}
library(autovarCore)

# simulate data
N = 100
depression <- rnorm(N)
activity <- rnorm(N)
activity_lag1 <- c(NA, activity[1:(N -1)])

depression <- depression + 0.5 * activity_lag1 
d <- data.frame(depression, activity)

models_found <- autovarCore::autovar(d, selected_column_names = c('activity', 'depression'))

# Show details for the best model found
summary(models_found[[1]]$varest$varresult$depression)
```

AutovarCore is a simplified version of a more extensive package ***autovar*** [@[@R-autovarCore], which was used in several publications [@VanderKrieke2015, @Emerencia2016]. Further information can be found on [http://autovar.nl](https://autovar.nl/) and [http://autovarcore.nl](https://autovarcore.nl/) 


### Package 'qgraph`
\index{Packages!qgraph}

Package ***qgraph*** [@Epskamp2012] can be used to fit, visualise and analyse graphical networks.

In the example below, qgraph is used to fit a network on the 'Critical Slowing Down' (CSD) data set, which is included in package 'emaph' (see ?csd and Chapter \@ref(networks)). 

```{r csd83, out.width = "100%", fig.asp = .4, message = FALSE, warning = FALSE, fig.cap="Network of mood items from CSD data set"}
library(qgraph)
library(emaph)

# get mood_ scores from csd data set
d <- subset(csd, 
            subset = phase == "exp: db: no change", 
            select = grep("mood_", names(csd))[1:5])


# Fit and plot regularized partial correlation network
g <- qgraph(cor_auto(d, detectOrdinal = FALSE),
       graph = "glasso", sampleSize = nrow(d),
       nodeNames = names(d),
       label.scale = FALSE, label.cex = .8, 
       legend = TRUE, legend.cex = .5,
       layout = "spring")
```


### Package 'bootnet'
\index{Packages!bootnet}

In the interpretation of fitted network plots, it is important to take the stability of the network into account. Intuitively, networks that are fit on small sample data sets will be less stable than networks based on lare data sets. One solution is to fit a large number of networks on subsets of the original data, through bootstrapping. In stable networks, the variance edges estimations will be small,  while in unstable networks, the variance will be high. This idea is implemented in package ***bootnet***  [@Epskamp2018a]. 

Below, the stability of the network that was fit in the prevous example is examined with bootnet: fifty networks are fit, based on fifty bootstrapped samples. In the results plot, the red line marks the strength of the edges in the full sample, while grey confidence intervals mark the distribution of the edge weights in the bootstraps. 

```{r out.width = "100%", fig.asp = .4}
library(bootnet)

g <- estimateNetwork(d, default = "EBICglasso",
                     corArgs = list(detectOrdinal = FALSE))
results <- bootnet(g, nBoots = 50, verbose = FALSE)
plot(results, order = "mean")
```


## Timeseries analysis

### Package 'lomb'
\index{Lomb-Scargle periodogram}

Disturbances in circadian rhythms have been related to depressive symptoms [see, e.g., @Saeb2015]. With so-called periodograms, these circadian rhytms can be detected in EMA data. Standard analysis techniques, however, expect regular timeseries, in which data are sampled at equidistant intervals. EMA data, typically, are not equidistant. One solution to this problem is to use the Lomb-Scargle periodogram procedure [@Lomb1976], which can be applied to unevenly-sampled timeseries as well. Package ***lomb*** [@ruf1999] implements this procedure.

```{r cs18f, out.width='98%'}
# code snippet 18.5: calculating a Lomb-Scargle periodogram
data(ibex, package = "lomb") 
lomb::lsp(ibex[2:3]) 
```
