# (PART) A Case Study {-}

# Early Warning Signs of Depression {#csd}

One of the promises of EMA is that it might detect signs of mental health
deterioration in an early stage. Subtle changes in time series of mood
variables, for example, might signal a depression relapse. If we can detect
these changes, preventive interventions can be triggered to avoid the relapse.

But what changes, exactly, should we look for? What are these early warning
signs?


## Critical Slowing Down
\index{Critical Slowing Down}

Critical Slowing Down (CSD) is a concept from dynamic systems theory. In dynamic
systems, state transitions are preceded by a change in which the system reacts
to disturbances. In a stable state, the system quickly recovers from
disturbances. Prior to a transition to a new state, however, the system takes
more and more time to recover back to its current state: the variance and
auto-correlation of the system increases [@Scheffer2009; @Dakos2008].

In this chapter, we re-analyze data from a study that aimed to demonstrate CSD
in EMA-data of a single patient with a history of major depression [@Groot2010,
@Kossakowski2017; @Wichers2016]. The patient, a 57-year old male, used EMA to
monitor himself during a 239-day single-case double-blind medication reduction
trial. During the experiment, he experienced a relapse, and Wichers and
colleagues showed how variance and auto-correlation in the EMA data increased,
several weeks prior to this relapse. The transition appeared to be preceded by
CSD.

We will try to reconstruct the finding, using an alternative analysis strategy.
One of the limitations of the Wichers et al analysis is that auto-correlation was
analyzed at lag 1 only (i.e., only the correlation between t and t-1 was
considered). With another analysis technique, called 'Detrended Fluctuation
Analysis', all lags can be considered.

To conduct the analysis, we need three R packages:
\index{tidyverse}
\index{nonlinearTseries}
\index{empah}

- Raw EMA data of this study were published in the public domain
[@Kossakowski2017]. We included the data in the `emaph` package.

- To manipulate the raw data and reconstruct the plots of the article, we are
going to use several functions from `tidyverse` packages.

- DFA is implemented in package `nonlinearTseries`, so we will need that as well.


```{r csd-lib, warning=FALSE}
# Required libraries for the CSD-study re-analysis.
library(emaph)
library(tidyverse)
library(nonlinearTseries)
```


## Plotting the Course of Depression 

Let's take a look at the development of the depressive symptoms of the patient
first. These were tapped with weekly assessments of the depression scale of the
'Symptom Checklist-90-Revised' [SCL-90-R; @Derogatis1994], a well-established
self-report questionnaire.

The code below reconstructs Figure 1 of the Wichers et al article. It plots the
SCL90-R depression scale score over the time. Around day 120, the depression
score increased considerable: the patient experienced a relapse.

```{r csd-scl90plot, out.width = "100%", fig.cap = "SCL-90 depression score, over the study period", fig.asp = .3}
# Plot depression score.
dep <- csd %>% select(dayno, scl90r_dep) %>%
  filter(!is.na(scl90r_dep)) %>% unique

# plot dep + change point (day 120 in our data)
ggplot(dep, aes(x = dayno, y = scl90r_dep, group = 1)) +
  geom_step() +
  xlab("Days") + ylab("SCL-90-R Depression") + 
  theme_classic()
```


## Mental state EMA Items

Wichers and colleagues selected 13 items from the full EMA data set, which they
grouped in 5 factors: positive affect (pa; 4 items), negative affect (`na`; 4
items), mental unrest (`mu`; 3 items), suspiciousness (`su`; 1 item), and worrying
(`wo`; 1 item). From these factors, an overall mental state sum score can be
calculated.

```{r csd-calc-ms}
# Mood states calculation.

# positive affect
pa_items <- c("mood_enthus", "mood_cheerf",
              "mood_strong", "mood_satisfi")

csd$pa <- csd %>%
  select(pa_items) %>%
  rowMeans(., na.rm = TRUE)
csd$pa <- -csd$pa

# negative affect
na_items <- c("mood_lonely", "mood_anxious",
              "mood_guilty", "mood_doubt")

csd$na <- csd %>%
  select(na_items) %>%
  rowMeans(., na.rm = TRUE)

# mental unrest
mu_items <- c("mood_irritat", "pat_restl",
              "pat_agitate")
csd$mu <- csd %>%
  select(mu_items) %>%
  rowMeans(., na.rm = TRUE)

# 'single-item' states
csd$su <- csd$mood_suspic
csd$wo <- csd$pat_worry

# global mental state score
csd$ms <- rowSums(csd[c("pa", "na", "mu", "su", "wo")])
```

Rows, in which one or more of the factors have missing values, are removed from
the analysis. In a full analysis, options to impute the missing values could and
should be considered. However, since only 3 of the 1476 rows have missing item
scores, not much is probably lost by running a simple complete-case analysis.

```{r csd-na}
# Missing values removal.

# count number of items with missing items, per row
csd$nna <- csd %>%
  select(matches("mood_")) %>%
  is.na(.) %>% rowSums

# drop rows with missing values
csd <- csd %>% filter(nna == 0)
```


## Running the DFA
\index{DFA analysis}

We are now ready to run the DFA analysis. We split the full series in 31-day
overlapping windows, in steps of 1 day (i.e., day 1-31, day 2-32, etc.),
calculate the DFA of each window and save that value so that we can compare it
to the weekly depression assessments.


```{r csd-dfa}
# Running the DFA.

# prepare result rows: one row for each day
d <- NULL
d <- csd %>% 
  group_by(dayno) %>%
  summarise(ms = mean(ms))
d$ms_dfa = NA

# determine DFA, in a moving window of 31 days
window <- 31
for (i in seq(window, max(csd$dayno), 1)) {

  # get the sliding window data
  w <- subset(csd, dayno > (i - window) & dayno <= i)

  # dfa: ms
  dfa.analysis <- dfa(time.series = w$ms,
                      npoints = 30,
                      window.size.range = c(10, nrow(w)),
                      do.plot = FALSE)

  fgn.estimation <- estimate(dfa.analysis,
                            do.plot = FALSE,
                            fit.col = "blue", fit.lwd = 2, fit.lty = 2,
                            main = "Fitting DFA to fGn")
  
  d$ms_dfa[d$dayno == i] <- fgn.estimation
}
```



## Results

We can now plot the DFA indicator over time, to see whether it peaks prior to
the onset of the relapse. As can be seen, the DFA-indicator clearly peaks prior
to the onset of the relapse, around day 110 (where the color changes to red).
Interestingly, a second peak is reached at the end of the experiment, around
day 239, indicating - perhaps - a recovery from the relapse.

```{r csd-dfaplot, eval = TRUE, out.width = "100%", fig.asp = .35, fig.cap = "Results of the DFA analysis.", warning = FALSE}
# Plot DFA results.
ggplot(na.omit(d),
      aes(x = dayno, y = ms_dfa, 
          colour = factor(dayno < 120), 
          group = 1)) +
  geom_point() +
  geom_line() + 
  ylab("dfa (31-day window)") + 
  xlim(c(0, 250)) + 
  guides(color = FALSE) + 
  theme_classic()
```


## Discussion

Our re-analysis replicates the main finding of the Wichers et al article
([@Wichers2016]): several weeks prior to a depression relapse, as predicted by
complex systems theory, the variance and auto-correlation in EMA mood ratings
increased.

Potential clinical applications, of course, are clear. If clinically relevant
changes can be predicted algorithmically from EMA ratings, automated patient
feedback systems could help to prevent pending deterioration or consolidate the
path towards recovery.

One swallow, however, does not make summer. Yes, in this case, the DFA-indicator
peaked prior to the relapse. This could be a mere coincidence. The predictive
power of CSD needs to be confirmed in larger samples and prospective studies.
Given the theoretical background, successful applications of CSD in other
domains, and the present finding, the spending of resources on such studies
seems warranted.

Important additional questions remain to be answered. When it predicts a state
change, what is the expected delay between this prediction and the change? Does
a critical DFA-value exist? Given that critical value, are false positive and
false negative rates of this prediction acceptable? These are important
questions that should be answered before any clinical application of DFA can be
considered.

Re-analysis of data from completed clinical studies, in which EMA data were
collected, may be a first step to further explore the value of CSD. One option,
for example, would be to re-analyze data from the E-COMPARED study
[@Kleiboer2016]. In this trial, patients, who were treated for major
depression, completed weekly self-report questionnaires [the Patient Health
Questionnaire; PHQ-8, @Kroenke2009] and daily EMA mood ratings throughout
treatment, which lasted up to 16-week. Since CSD is an indicator of *any* state
change (i.e., irrespective of whether the change is clinically "good" or "bad"),
theory would predict a (lagged) relationship between CSD (i.e., the DFA-score)
and clinically significant changes in PHQ-scores [@Jacobson1991].






