
# (PART) Analytic Approaches {-}

# Feature Extraction {#features}

EMA data are streams: time-series that represent processes which may span hours,
days, weeks or even months. To the beginning EMA researcher, it may not be
immediately clear how to deal with these data. How, for example, can these
series be summarized to study differences between individuals or changes within
individuals over time? What 'qualities' or 'features' of a time-series should be
considered? What are the available options? In this chapter, we will discuss
some of the features that can be extracted from EMA time-series.


## Simulated EMA Time-series Example
\index{Simulating data} 

We will focus on a simulated three-week time-series of EMA mood responses of a
single person, in which ratings were collected five times per day. Figure
\@ref(fig:feat-plot) shows a plot of the simulated scores.

Stare at the figure for a moment. How would you characterize the development of
scores over time? What features would you want to extract? How would you
quantify these features?

```{r feat-sim, echo = FALSE}
## Constructing the simulated EMA time-series

# libraries
library(ggplot2)
library(gridExtra)
library(psych)

# simulate the signal -------
set.seed(123)
n_measurements_per_day <- 5 
n_days <- 21
n <- n_measurements_per_day * (n_days)
m <- 3
sdev_min <- 1
sdev_max <- 4
trend  <- 3
missingness_prob = c(0.1, 0.5)
autocorrelation = .6

# time
t <- seq(0, (n - 1) * 1 /5, by =  1/5)

# signal
s <- numeric(length(t))

# periodicity
s <- seq(sdev_min, sdev_max, length.out = n) * sin(1   * t * 2 * pi)
s <- s + 0.5 * seq(sdev_min, sdev_max, length.out = n) * sin(1/7 * t * 2 * pi)

# mean + trend + sd (heteroscedastic)
s <- s + rnorm(n, 
               mean = seq(m, m + trend, length.out = n), 
               sd = 0.5)


s <- s + 1 * arima.sim(model = list(ar = autocorrelation), n = n)

# missingness
m1 <- rbinom(n, 1, seq(missingness_prob[1], missingness_prob[2], length.out = n))
s[m1 == 1] <- NA 

# in range
s[s > 10] <- 10
s[s < 0] <- 0

# combine in data.frame
d <- data.frame(t, s)
```

```{r feat-plot, echo = FALSE, out.width = "100%", fig.asp = .5, fig.cap = "A simulated three-week time-series of EMA mood ratings."}
# plot signal
e <- subset(d, !is.na(s))
ggplot(e, aes(x = t, y = s)) + 
  geom_line() + 
  geom_point() +
  ylab("mood") + xlab("days") +
  ylim(0, 10) + 
  theme_bw() + theme(panel.grid.minor = element_blank())
```

## Central Tendency and Variability

Features do not necessarily have to be complex. Familiar measures of central
tendency of EMA time-series, such as the mode or the mean, can be useful
predictors of traditional clinical assessments. Standard measures of the
variability of EMA scores, such as the standard deviation (SD), have been shown
to have diagnostic value [see, e.g., @bowen2006]. You should not hesitate to
consider these statistics when they help you to answer your research question.

\index{Stationarity}
You should check, though, whether the implicit assumptions behind the statistics
are correct. When you use the mean and the standard deviation to summarize a
time-series, you assume that the data-generating process behind the series is
stable over time - in technical terms: that this process is *stationary* 
[@wikiStationary]. If this assumption is incorrect (i.e., if the
process is unstable or non-stationary), the mean and SD are biased statistics.

Our simulated EMA time-series is non-stationary, as can be seen in Figure
\@ref(fig:feat-plot-mean). We observe a trend: the mood of the person increases
over time. EMA scores tend to be below the mean in the beginning, and above the
mean at the end. The overall mean (*M* = `r round(mean(e$s), 1)`) overestimates
the EMA scores of the first week (*M* = `r round(mean(e$s[e$t < 8]), 1)`), and
underestimates the scores of the last week (*M* =
`r round(mean(e$s[e$t > 13]), 1)`). If you use the overall mean as a feature of
the time-series in your analyses, you ignore the mood improvements over the
three-week period. In clinical research, your would probably not want to miss
this other important feature.

When a trend exists in a series, the overall standard deviation is
overestimated, because of the extra variability that is introduced by the
changes of the mean. This can be checked by comparing the SD of the full series
to the SD of smaller parts. In our example, the overall SD is
`r round(sd(e$s), 1)`. If we calculate the average of the SD of scores of each
day, however, the estimated SD is
`r round(mean(aggregate(e$s, by = list(day = floor(e$t)), FUN = sd)$x), 1)`. If
the additional variability introduced by the trend is cancelled out, our
estimate of variability decreases considerably. 

```{r feat-plot-mean, echo = FALSE, out.width = "100%", fig.asp = .5, fig.cap = "EMA time-series, with reference lines for the mean (red line) and the mean +/- 1 standard deviation range (the area between the two blue lines). Both statistics are informative, but obviously do not do full justice to the variation in observations over time."}
# plot signal with mean and variance
e <- subset(d, !is.na(s))
ggplot(e, aes(x = t, y = s)) +
  geom_hline(yintercept = mean(e$s, na.rm = TRUE), color = "red") + 
  geom_hline(yintercept = mean(e$s, na.rm = TRUE) + sd(e$s, na.rm = TRUE), color = "blue", linetype = 2) +
  geom_hline(yintercept = mean(e$s, na.rm = TRUE) - sd(e$s, na.rm = TRUE), color = "blue", linetype = 2) +
  #geom_line() + 
  geom_point() + 
  ylab("mood") + xlab("days") +
  ylim(0, 10) +
  theme_bw() + theme(panel.grid.minor = element_blank())
```

## Modelling the Trend
```{r feat-lm, echo=FALSE, include=FALSE}
fm = lm(s ~ t, e)
```

Now that we know that the overall mean and SD are sub-optimal features of this
time-series, the question is how we can do better. What other features can we
extract? One option would be to characterize the series in simple regression
terms, via an intercept and a slope. By doing so, we can economically describe
the dynamics of the series with two numbers. As can be seen in Figure
\@ref(fig:feat-plot-lm), the regression model provides a more realistic and informative summary. At the start, the estimated mean - the intercept - is
`r round(coef(fm)[1], 1)`. This mean increases approximately
`r round(coef(fm)[2] * 7,  1)` points per week (the slope), to a final estimated
mean of `r round(coef(fm)[1], 1) + round(coef(fm)[2] * 21, 1)`. Comparing Figure
\@ref(fig:feat-plot-mean) and Figure \@ref(fig:feat-plot-lm), we see that
Figure \@ref(fig:feat-plot-lm) is more successful in capturing the variation in
the series.

```{r feat-plot-lm, echo = FALSE, out.width = "100%", fig.asp = .3, fig.cap = "EMA time-series, with a regression reference line (red) and the residual error SD range around this line (the area between the two blue lines)"}
ggplot(e, aes(x = t, y = s)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  geom_abline(intercept = coef(fm)[1] + sd(resid(fm)), slope = coef(fm)[2], color = "blue", linetype = 2) + 
  geom_abline(intercept = coef(fm)[1] - sd(resid(fm)), slope = coef(fm)[2], color = "blue", linetype = 2) + 
  geom_point() + 
  ylab("mood") + xlab("days") +
  ylim(0, 10) +
  theme_bw() + theme(panel.grid.minor = element_blank())
```

By using regression, we can also find a better estimate of the SD. Remember how
the SD is conceptually defined as the average distance between the scores and
the mean. Since the mean is modeled by the regression line, we can estimate the
SD by calculating the standard deviation of the residuals of the regression
model that we used to retrieve the intercept and slope. For our series, this
results in an estimated SD of `r round(sd(resid(fm)), 1)`. The new SD estimate
is smaller than the overall SD, as expected. By accounting for the extra
variability introduced by the trend, we get a more accurate estimate of the
variability around the mean at each point in time. However, the new estimate is
still higher than the daily estimate that we calculated above. Figure
\@ref(fig:feat-plot-resid) provides a first indication of what is going on. In
the figure, absolution regression residuals are plotted over time, with a
regression line superposed. This reveals that residuals increase as time goes
by. We discovered a new feature of the series: heteroscedasticity. 

```{r feat-plot-resid, echo = FALSE, out.width = "100%", fig.asp = .3, fig.cap = "Plot of absolute regression residuals over time, revealing heteroscedasticity."}
e$residuals <- rstandard(fm)
e$week <- (e$t %/% 1 )+ 1

ggplot(subset(e, week < 20), aes(x = t, y = abs(residuals))) +
  geom_point() + 
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ylab("absolute residuals") + xlab("week") +
  theme_bw() + theme(panel.grid.minor = element_blank())
```

## Missing Values
\index{Missing values}

```{r feat-plot-na-fm, echo = FALSE, include = FALSE}
d$t_g <- cut(t, 21, labels = 1:21)
p <- prop.table(table(d$t_g, is.na(d$s)), 1)
p <- as.data.frame(p)
p <- subset(p, Var2 == TRUE)
fm2 <- lm(p$Freq ~ as.numeric(p$Var1))
```

In EMA research, you should be prepared to deal with missing values. Especially
with active EMA, in which participants have to consciously respond to prompted
questions several times a day, non-response is inevitable. Missing values are
typically considered to be a nuisance rather than a feature of EMA time-series.
Non-response, however, is an important characteristic, that may even have
clinical relevance. Missing values are also present in our simulated EMA
time-series example, as you may have found out already when you inspected Figure
\@ref(fig:feat-plot). The points, denoting the observed responses, are denser in
the beginning of the series. Over the course of the study period, the
probability of missed ratings increases. This becomes immediately clear from
Figure \@ref(fig:feat-plot-na), in which the percentage of missed ratings is
plotted per day. Every week, the probability of missed ratings increases
approximately `r round(coef(fm2)[2]*7*100)`%.

```{r feat-plot-na, echo = FALSE, out.width = "100%", fig.asp = .3, fig.cap = "Percentage of missed mood ratings, per day, over the three-week study period, with a regression line (red) superposed, revealing a familiar trend in EMA data."}
ggplot(p, aes(x = as.numeric(Var1), y = Freq * 100)) + 
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  geom_point() + 
  ylab("missed ratings (%)") + xlab("days") +
  ylim(0, 100) + 
  theme_bw() + theme(panel.grid.minor = element_blank())
```

Now that we now about these missing values, the question arises how to deal with
them? You can choose to ignore the missing values (assuming that the occurrence
of missing values was driven by a completely random process). Alternatively, you
could try to replace the missing values with plausible values (in which case it
becomes important to think about the process that drives the missingness). If
you decide to impute, there are several options. For example, you can replace
missing values with the mean, the last-known value, an interpolated value or
with a smoothing technique such as the Kalman-filter [see @hoogendoorn2017, for
a discussion]. Figure \@ref(fig:feat-plot-interpolation) illustrates the
linear interpolation approach: missing values (marked by red dots) are replaced
by values that lie between the non-missings. Imagine that we would have used the
mean to impute. Can you see why that would have disrupted local patterns in the
series?

```{r feat-plot-interpolation, echo = FALSE, out.width = "100%", fig.asp = .3, fig.cap = "EMA time series, with missing values (red) imputed through interpolation."}
# Imputation through interpolation 
impute_interpolation <- function(x) {
  require(zoo)
  
  y = zoo(x)
  
  # fill initial and trailing NA
  if (is.na(y[1])) y[1] = y[which.max(!is.na(y))]
  if (is.na(y[length(y)])) y[length(y)] = y[max((!is.na(y))*(1:length(y)))]
  
  # interpolate
  y_ = as.numeric(na.approx(y))
  
  y_
}

d$s_imputed <- impute_interpolation(d$s)
ggplot(d, aes(x = t, y = s_imputed)) +
  geom_line(color = 1) + 
  geom_point(aes(color = is.na(s))) + 
  scale_color_manual(values=c("black", "red")) +
  guides(color = FALSE) + 
  theme_bw() + theme(panel.grid.minor = element_blank())
```


## Auto-correlation
\index{Auto-correlation}

When a series is correlated with delayed copies of itself, we say that it is
auto-correlated. In repeated measurements of natural phenomena, this
auto-correlation can often be found. The temperature of today is correlated with
the temperature of recent days. Similarly, EMA mood ratings, taken at time t,
are typically correlated with ratings taken at t-1, t-2, etc. Studying the
auto-correlation of a time-series at different delays can be very revealing, as
illustrated by the auto-correlation plot in Figure
\@ref(fig:feat-plot-autocorr), in which the correlation of the EMA series with
itself is plotted, for various delays (or 'lags' as these delays are called).
First, the plot reveals that observations are correlated with previous values at
lag 1: this series is auto-correlated. Second, there appears to be a pattern in
the correlations at later lags: positive and negative correlations alternate, in
a pattern that seems to reflect periodicity.

```{r feat-plot-autocorr, echo = FALSE, out.width = "100%", fig.asp = .3, fig.cap = "Autocorrelation plot of the EMA mood series, revealing periodicity."}
op = par(mar = c(4, 4, 1, 1))
acf(e$s, type = "partial", lag.max = 35, main = "", ylab = "corr", xlab = "lag")
par(op)
```

### Rolling Statistics

From the auto-correlation analysis, we learned that describing the EMA series as
a simple trend might be too simplistic. The mean does not increase in a simple
linear fashion. There seem to be periodic components in the series as well: the
signal appears also to be characterized by a series of increases and decreases
in the mean mood level. In the plot of the raw series, this is not immediately
clear. However, if we "smooth" the series, by calculating the mean as it
develops over time, the periodicity in the series is becoming more clear. Figure
\@ref(fig:feat-plot-roll) illustrates this.

```{r feat-plot-roll, echo = FALSE, out.width = "100%", fig.asp = .3, fig.cap = "EMA mood series, with a rolling mean superposed."}
d$rolling_mean <-
  as.numeric(
    zoo::rollapply(
      as.numeric(d$s),
      width = 5,
      FUN = function(x)
        mean(x, na.rm = TRUE),
      fill = NA,
      align = "right",
      partial = TRUE
    )
  )

e <- subset(d, !is.na(s))
ggplot(e, aes(x = t, y = rolling_mean)) +
  geom_point(aes(y = s), color = "grey", size = .8) + 
  geom_line(size = 1.2) + 
  ylab("Mood (rolling mean)") + xlab("days") +
  ylim(0, 10) + 
  theme_bw() + theme(panel.grid.minor = element_blank())

```


## Periodicity
\index{Circadian rhythms}

Both the auto-correlation analysis and the moving average plot suggest
periodicity in the EMA time series. Suppose we want to know more about this
periodicity. Is there a way to quantify it? Yes, there is. With a technique
called Fourier Analysis, the strength ('power') of various frequencies can
be quantified. Figure \@ref(fig:feat-plot-pow) shows what happens if we run a Fourier analysis of our time series (with R's built-in function `spec.pgram`).

```{r feat-plot-pow, echo = FALSE, out.width = "100%", fig.asp = .4, fig.cap = "periodogram of the EMA time series, revealing a one-day and a one-week period."}
fa <- spec.pgram(ts(impute_interpolation(d$s), frequency = n_measurements_per_day), 
                 detrend = TRUE, demean = TRUE, 
                 main = "periodogram",  sub = "",
                 xlab = "frequency", log = "no",
                 ylab = "power", type = "h",
                 spans = NULL, taper = 0.01, ci = FALSE)
```

There are two peaks in the figure: at a frequency of 1 (day) and at a frequency
of `r round(fa$freq[order(fa$spec, decreasing = TRUE)][1], 2)` (about one week).
This series is *circadian*: mood ratings follow a daily and a weekly
pattern. This is a common feature of EMA data, because of the circadian
rhythms in our body and in our environment (our biological clocks, day-night
cycles, seasons) [@Doherty2018; @Frank2000; @Frank2007; @Karatsoreos2014; @VanSomeren2000;
@Tahmasian2013].


## Discussion

In this chapter, we used a number of statistical indicators to characterize a
single time-series. By extracting these indicators (or "features"), we
identified several regularities. We learned that the mean, variance and
missingness increased over time, and we identified clear signs of circadian
rhythmicity. You might have identified some of these features when you first
inspected Figure \@ref(fig:feat-plot). The increase in the mean was pretty easy
to spot. However, the systematic increases in the variance and the missingness
were less clear. Likewise, although you might have spotted the periodicity in
the series, the circadian components were probably too subtle to spot for the
naked eye.

The aim of the chapter was to illustrate the options available to the EMA
researcher to quantitatively summarize EMA series. Other options, of course,
exist. In chapter \@ref(csd), for example, we will see how subtle changes in
variation and auto-correlation in EMA mood scores can be summarized in a single
dynamic statistical feature that may be predictive indicator of significant
future state changes.

Finally, you might have been disappointed by the lack of code examples in this
chapter. We did not include them, to avoid unnecessary distraction from the
conceptual explanation. Instead, we provide a full code listing below, to allow
you learn how to apply the feature extraction techniques yourself. 

\index{Simulating data} 

```{r feat-allcode, echo = TRUE, eval = FALSE}
## (De)constructing a simulated EMA time-series

# libraries ----------------
library(ggplot2)
library(gridExtra)
library(psych)
library(zoo)

# simulate the signal -------
set.seed(123)
n_measurements_per_day <- 5
n_days <- 21
n <- n_measurements_per_day * (n_days)
m <- 3
sdev_min <- 1
sdev_max <- 4
trend  <- 3
missingness_prob = c(0.1, 0.5)
autocorrelation = .6

## time
t <- seq(0, (n - 1) * 1 / 5, by =  1 / 5)

## signal
s <- numeric(length(t))

## periodicity
s <- seq(sdev_min, sdev_max, length.out = n) * sin(1   * t * 2 * pi)
s <-
  s + 0.5 * seq(sdev_min, sdev_max, length.out = n) * sin(1 / 7 * t * 2 * pi)

## mean + trend + sd (heteroscedastic)
s <- s + rnorm(n,
               mean = seq(m, m + trend, length.out = n),
               sd = 0.5)

#3 auto-correlation
s <- s + 1 * arima.sim(model = list(ar = autocorrelation), n = n)

## missingness
m1 <-
  rbinom(n, 1, seq(missingness_prob[1], missingness_prob[2], length.out = n))
s[m1 == 1] <- NA

## in range
s[s > 10] <- 10
s[s < 0] <- 0

## combine in data.frame
d <- data.frame(t, s)

# plot -------------------------
e <- subset(d,!is.na(s))

# raw series 
ggplot(e, aes(x = t, y = s)) +
  geom_line() +
  geom_point() +
  ylab("mood") + xlab("days") +
  ylim(0, 10) +
  theme_bw() + theme(panel.grid.minor = element_blank())


# plot mean and variance --------
ggplot(e, aes(x = t, y = s)) +
  geom_hline(yintercept = mean(e$s, na.rm = TRUE), color = "red") +
  geom_hline(
    yintercept = mean(e$s, na.rm = TRUE) + sd(e$s, na.rm = TRUE),
    color = "blue",
    linetype = 2
  ) +
  geom_hline(
    yintercept = mean(e$s, na.rm = TRUE) - sd(e$s, na.rm = TRUE),
    color = "blue",
    linetype = 2
  ) +
  #geom_line() +
  geom_point() +
  ylab("mood") + xlab("days") +
  ylim(0, 10) +
  theme_bw() + theme(panel.grid.minor = element_blank())


## The trends ---------------------

## run a linear regression of mood ratings by time
fm = lm(s ~ t, e)

## mean
ggplot(e, aes(x = t, y = s)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  geom_abline(
    intercept = coef(fm)[1] + sd(resid(fm)),
    slope = coef(fm)[2],
    color = "blue",
    linetype = 2
  ) +
  geom_abline(
    intercept = coef(fm)[1] - sd(resid(fm)),
    slope = coef(fm)[2],
    color = "blue",
    linetype = 2
  ) +
  geom_point() +
  ylab("mood") + xlab("days") +
  ylim(0, 10) +
  theme_bw() + theme(panel.grid.minor = element_blank())


## residuals
e$residuals <- rstandard(fm)
e$week <- (e$t %/% 1) + 1

ggplot(subset(e, week < 20), aes(x = t, y = abs(residuals))) +
  geom_point() +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  ylab("absolute residuals") + xlab("week") +
  theme_bw() + theme(panel.grid.minor = element_blank())


# Missing values ----------------------

d$t_g <- cut(t, 21, labels = 1:21)
p <- prop.table(table(d$t_g, is.na(d$s)), 1)
p <- as.data.frame(p)
p <- subset(p, Var2 == TRUE)
fm2 <- lm(p$Freq ~ as.numeric(p$Var1))

ggplot(p, aes(x = as.numeric(Var1), y = Freq * 100)) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  geom_point() +
  ylab("missed ratings (%)") + xlab("days") +
  ylim(0, 100) +
  theme_bw() + theme(panel.grid.minor = element_blank())


## imputation through interpolation
impute_interpolation <- function(x) {
  require(zoo)
  
  y = zoo(x)
  
  # fill initial and trailing NA
  if (is.na(y[1]))
    y[1] = y[which.max(!is.na(y))]
  if (is.na(y[length(y)]))
    y[length(y)] = y[max((!is.na(y)) * (1:length(y)))]
  
  # interpolate
  y_ = as.numeric(na.approx(y))
  
  y_
}


d$s_imputed <- impute_interpolation(d$s)
ggplot(d, aes(x = t, y = s_imputed)) +
  geom_line(color = 1) +
  geom_point(aes(color = is.na(s))) +
  scale_color_manual(values = c("black", "red")) +
  guides(color = FALSE) +
  theme_bw() + theme(panel.grid.minor = element_blank())


# Autocorrelation ---------------------------
acf(e$s,
    type = "partial",
    lag.max = 35,
    main = "")


# Rolling mean ------------------------------
d$rolling_mean <-
  as.numeric(
    zoo::rollapply(
      as.numeric(d$s),
      width = 5,
      FUN = function(x)
        mean(x, na.rm = TRUE),
      fill = NA,
      align = "right",
      partial = TRUE
    )
  )

e <- subset(d,!is.na(s))
ggplot(e, aes(x = t, y = rolling_mean)) +
  geom_point(aes(y = s), color = "grey", size = .8) +
  geom_line(size = 1.2) +
  ylab("Mood (rolling mean)") + xlab("days") +
  ylim(0, 10) +
  theme_bw() + theme(panel.grid.minor = element_blank())


# Frequency analysis --------------------------

fa <-
  spec.pgram(
    ts(impute_interpolation(d$s), frequency = n_measurements_per_day),
    detrend = TRUE,
    demean = TRUE,
    main = "periodogram",
    sub = "",
    xlab = "frequency",
    log = "no",
    ylab = "power",
    type = "h",
    spans = NULL,
    taper = 0.01,
    ci = FALSE
  )
```